A forest is more than its trees: haplotypes and inferred ARGs

### 2024-10-11 paper/writeup.tex
citations

\comment{idea: look at ratio of unary-to-not over time; does it change with
    selection? or modes of recombination?}

@@

There has been substantial recent progress
in the problem of ``ARG inference'',
which seeks to infer (portions of) the ``ancestral recombination graph'' (or, ARG)
that describes how a set of genotypes samples are related to each other
at each position of the genome \citep{XXX}.
One way of viewing the ARG is as a sequence of ``marginal trees'',
i.e., the genealogical trees that describe how each portion of the genome
was inherited by the focal genomes.
This is reflected in methodology of some ARG inference methods \citep[e.g.,][]{Relate},
in metrics used to assess inference accuracy \citep{XXX},
as well as in basic terminology.
For instance, 
the ``succinct tree sequence'',
introduced by \citet{kelleher2016efficient},
is a common format for describing these inferred ARGs,
and is seeing wide use thanks in part to its efficiency and accompanying reliable toolkit,
\tskit \citep{XXX}.

@@

Haplotype identity has been relatively overlooked in the literature on ARG inference so far --
all metrics that have been used so far to measure accuracy of inferred ARGs
depend only on the sequence of marginal trees,
not on sharing of ancestral haplotype identities between them.
Suppose two methods are used to infer ARGs from the same set of genotyped samples.
Although both ARGs describes relationships between many haplotypes,
only those haplotypes that represent the focal set of genotyped samples
can be unambiguously said to represent the same thing in both.
So, most metrics only describe these relationships,
which ignores the additional information carried by the structure of other haplotypes.
For instance, \citet{XXX} compared true and inferred ARGs
using average XXX distance between trees across a regular sequence of genomic positions,
using sampled genotypes as labels,
while \citet{brandt2022evaluation} compared times to most recent common ancestor
between pairs of sampled genomes.
Neither is affected by shared haplotype structure -
two ARGs could be identiical by either measure
but have completely different identity by descent (IBD) segment structure
(as defined in the context of the ARG, \citep{XXX}).

@@

It is helpful to first describe what we compute in the simple case.
\figref{fig:node-spans}A shows the spans for each node in three tree sequences:
for each node, the top point is the total span in a ``true'' tree sequence
that was simulated with \texttt{msprime} \citep{msprime}
in a way that included spans over which nodes are unary in the output
\comment{(using which option?)}.
The bottom point is the span of the node after \emph{simplifying} the tree sequence
\citep{kelleher2018efficient},
which removes all unary nodes from all trees.
\comment{Usually (?)} between the two is the span of the node after applying \algorithmref{alg:extend}
to the simplified tree sequence.
So, the difference between the middle and bottom points is the amount by which
the span of that node has been extended.
\figref{fig:node-spans}B \comment{TODO} shows how much of this span is incorrect:
for each node, it gives the percent of the span after running \algorithmref{alg:extend}
that is incorrect (i.e., not present in the original tree sequence).
The values in these plots (perhaps summed over nodes)
therefore quantify how much the ancestral haplotypes represented by each node
were correctly, and incorrectly, extended.
They also tell us how much of the relationships in the original tree sequence
are represented in the result.

@@

Other authors \citep{XXX} have used
an average Robinson-Foulds distance \citep{robinson1981comparison}
as a measure of disagreement between tree sequences.
This could be computed in a very similar way:
instead of $m(n, \alpha(n))$ define
\begin{align*}
    m'(n, \T_2) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{\exists n_2: s(T^{(2)}_k, n_2) = s(T^{(1)}_k, n)} ,
\end{align*}
the total span over which there is \emph{some} node in $\T_2$ whose set of descendant samples
matches $s(T_1,n)$.
Then the average Robinson-Foulds distance (averaged over locations in the genome)
is
\begin{align*}
    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
\end{align*}
\comment{Check this!}
In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
across all trees, but average Robinson-Foulds distance allows a different node to match
on each tree.
The other differences are that \emph{average} Rbonson-Foulds distance
normalizes by sequence length, and is symmetrized.
Robinson-Foulds distance between two trees
was defined by \citet{robinson1981comparison}
to the minimum number of edge contraction/expansion operations needed to move
from one tree to the other
(which they then show is equal to the number of edges that induce different splits on the labels).
It would be interesting to define a similar metric on tree sequences,
perhaps using the Subgraph-Prune-and-Regraft moves of \citet{singer}.

@@

In this paper, we have taken a ``historical'' view of the ARG --
i.e., that each aspect of an inferred ARG is intended to represent
a portion of some particular historical genome
(for instance, the MRCA of some set of sampled genomes).
Furthermore, \figref{fig:conceptual}A\&B implicitly takes the position
that relationships \emph{not} depicted in the ARG are implied to not exist.
This is slighly disingenuous -- an alternative and more accurate interpetation
of the ARG of \figref{fig:conceptual}A\&B would be that we do not know
how node 2 inherited from node 4 on the right-hand interval,
rather than saying that it specifically did not pass through node 3.
The ``simplification'' algorithm of \citet{XXX},
or the Hudson algorithm for coalescent simulation
as implemented in msprime \citep{kelleher2016efficient},
each specifically discard information about any such ``non-coalescent'' nodes;
so the correct interpretation of these ARGs is a lack of knowledge.
In this paper, we have shown that this information can be largely replaced.

>>>>

@@

There has been substantial recent progress
in the problem of ``ARG inference'',
which seeks to infer (portions of) the ``ancestral recombination graph'' (or, ARG)
that describes how a set of genotypes samples are related to each other
at each position of the genome \citep{speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust,lewanski2024introduction}.
One way of viewing the ARG is as a sequence of ``marginal trees'',
i.e., the genealogical trees that describe how each portion of the genome
was inherited by the focal genomes.
This is reflected in methodology of some ARG inference methods \citep[e.g., Relate][]{speidel2019method},
in metrics used to assess inference accuracy \citep{XXX},
as well as in basic terminology.
For instance, 
the ``succinct tree sequence'',
introduced by \citet{kelleher2016efficient},
is a common format for describing these inferred ARGs,
and is seeing wide use thanks in part to its efficiency and accompanying reliable toolkit,
\tskit \citep{XXX}.

@@

Haplotype identity has been relatively overlooked in the literature on ARG inference so far --
all metrics that have been used so far to measure accuracy of inferred ARGs
depend only on the sequence of marginal trees,
not on sharing of ancestral haplotype identities between them.
Suppose two methods are used to infer ARGs from the same set of genotyped samples.
Although both ARGs describes relationships between many haplotypes,
only those haplotypes that represent the focal set of genotyped samples
can be unambiguously said to represent the same thing in both.
So, most metrics only describe these relationships,
which ignores the additional information carried by the structure of other haplotypes.
For instance, \citet{kelleher2019inferring} and \citet{zhang2023biobankscale}
compared true and inferred ARGs
using average Robinson-Foulds and Kendall-Colijn distances between trees
across a regular sequence of genomic positions,
using sampled genotypes as labels,
while \citet{brandt2022evaluation} compared times to most recent common ancestor
between pairs of sampled genomes.
Neither is affected by shared haplotype structure -
two ARGs could be identiical by either measure
but have completely different identity by descent (IBD) segment structure
(as defined in the context of the ARG, \citep{XXX}).
\citet{deng2021distribution} evaluated agreement of distributions of
distances along the genome between tree topology changes,
and \citet{zhang2023biobankscale} defined a generalization of Robinson-Foulds distance
that is the total variation distance between the induced distribution on genotypes;
neither of which measure sharing of haplotypes between adjacent trees.

@@

It is helpful to first describe what we compute in the simple case.
\figref{fig:node-spans}A shows the spans for each node in three tree sequences:
for each node, the top point is the total span in a ``true'' tree sequence
that was simulated with \texttt{msprime} \citep{kelleher2016efficient}
in a way that included spans over which nodes are unary in the output
\comment{(using which option?)}.
The bottom point is the span of the node after \emph{simplifying} the tree sequence
\citep{kelleher2018efficient},
which removes all unary nodes from all trees.
\comment{Usually (?)} between the two is the span of the node after applying \algorithmref{alg:extend}
to the simplified tree sequence.
So, the difference between the middle and bottom points is the amount by which
the span of that node has been extended.
\figref{fig:node-spans}B \comment{TODO} shows how much of this span is incorrect:
for each node, it gives the percent of the span after running \algorithmref{alg:extend}
that is incorrect (i.e., not present in the original tree sequence).
The values in these plots (perhaps summed over nodes)
therefore quantify how much the ancestral haplotypes represented by each node
were correctly, and incorrectly, extended.
They also tell us how much of the relationships in the original tree sequence
are represented in the result.

@@

Other authors \citep[e.g.,]{kelleher2019inferring} have used
an average Robinson-Foulds distance \citep{robinson1981comparison}
as a measure of disagreement between tree sequences.
This could be computed in a very similar way:
instead of $m(n, \alpha(n))$ define
\begin{align*}
    m'(n, \T_2) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{\exists n_2: s(T^{(2)}_k, n_2) = s(T^{(1)}_k, n)} ,
\end{align*}
the total span over which there is \emph{some} node in $\T_2$ whose set of descendant samples
matches $s(T_1,n)$.
Then the average Robinson-Foulds distance (averaged over locations in the genome)
is
\begin{align*}
    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
\end{align*}
\comment{Check this!}
In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
across all trees, but average Robinson-Foulds distance allows a different node to match
on each tree.
The other differences are that \emph{average} Rbonson-Foulds distance
normalizes by sequence length, and is symmetrized.
Robinson-Foulds distance between two trees
was defined by \citet{robinson1981comparison}
to the minimum number of edge contraction/expansion operations needed to move
from one tree to the other
(which they then show is equal to the number of edges that induce different splits on the labels).
It would be interesting to define a similar metric on tree sequences,
perhaps using the Subgraph-Prune-and-Regraft moves used by Singer \citet{deng2024robust}.

@@

In this paper, we have taken a ``historical'' view of the ARG --
i.e., that each aspect of an inferred ARG is intended to represent
a portion of some particular historical genome
(for instance, the MRCA of some set of sampled genomes).
Furthermore, \figref{fig:conceptual}A\&B implicitly takes the position
that relationships \emph{not} depicted in the ARG are implied to not exist.
This is slighly disingenuous -- an alternative and more accurate interpetation
of the ARG of \figref{fig:conceptual}A\&B would be that we do not know
how node 2 inherited from node 4 on the right-hand interval,
rather than saying that it specifically did not pass through node 3.
The ``simplification'' algorithm of \citet{kelleher2018efficient},
or the Hudson algorithm for coalescent simulation
as implemented in msprime \citep{kelleher2016efficient},
each specifically discard information about any such ``non-coalescent'' nodes;
so the correct interpretation of these ARGs is a lack of knowledge.
In this paper, we have shown that this information can be largely replaced.
